---
title: "Interpolation in Small Samples"
author: "Max Alletsee"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interpolation in Small Samples}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Sparse price information can become a problem when running Price Sensitivity Meter Analyses, particularly in small samples. In order to deal with this issue, the *pricesensitivitymeter* package has an option *interpolate* to approximate values on the price curves for all values where the actual information from respondents is missing.

This vignette illustrates the underlying problem and explains how the *interpolate* argument provides a solution.

## Illustrating the Problem

Consider an example with the following price information.

```{r}
set.seed(1976)

# create random variables
toocheap <- sort(round(rnorm(n = 50, mean = 8, sd = 1.5), digits = 2))
tooexpensive <- sort(round(rnorm(n = 50, mean = 20, sd = 1), digits = 2))
cheap <- sort(round(rnorm(n = 50, mean = 15, sd = 3), digits = 2))
expensive <- sort(round(rnorm(n = 50, mean = 16, sd = 3), digits = 2))

# setting up alternative "cheap"/"expensive" variables with a price step
pricestep_cheap <- which(cheap > 14.25 & cheap < 17.5)
cheap_alt <- cheap
cheap_alt[pricestep_cheap] <- round(mean(cheap_alt[pricestep_cheap]), digits = 2)

pricestep_expensive <- which(expensive > 14 & expensive < 16)
expensive_alt <- expensive
expensive_alt[pricestep_expensive] <- round(mean(expensive_alt[pricestep_expensive]), digits = 2)
```

When we run a Price Sensitivity Meter Analysis on the data and visualize it, we get the following result. (The plot is slightly simplified: It does not show the "too cheap" and "too expensive curves", as they are not relevant for the example. Moreover, it is slightly zoomed in.)

```{r, fig.width = 7, fig.height = 4.5}
library(pricesensitivitymeter)

# Running the analysis with price steps and without interpolation
psm_steps <- psm_analysis(toocheap = toocheap,
                          cheap = cheap_alt,
                          expensive = expensive_alt,
                          tooexpensive = tooexpensive,
                          validate = TRUE,
                          interpolate = FALSE)

library(ggplot2)

# Plot with price steps
psmplot_steps <-  ggplot(data = psm_steps$data_vanwestendorp, aes(x = price)) +
  geom_line(aes(y = ecdf_not_cheap, # line: not cheap
                colour = "not cheap"),
            size = 1) +
  geom_line(aes(y = ecdf_not_expensive, # line: not expensive
                colour = "not expensive"),
            size = 1) +
  geom_rug() +
  annotate(geom = "point", # Indifference Price Point (intersection of "not cheap" and "not expensive")
           x = psm_steps$idp,
           y = psm_steps$data_vanwestendorp$ecdf_not_expensive[psm_steps$data_vanwestendorp$price == psm_steps$idp],
           size = 5,
           shape = 18,
           colour = "#D55E00")

psmplot_steps +
  labs(x = "Price",
       y = "Share of Respondents (0-1)",
       title = "Price Sensitivity Meter Plot (without interpolation)")  +
  scale_colour_manual(name = "Legend",
                      values = c("not cheap" = "#009E73",
                                 "not expensive" = "#D55E00")) +
  annotate(geom = "text", # Label of Indifference Price Point
           x = psm_steps$idp + 1,
           y = psm_steps$data_vanwestendorp$ecdf_not_expensive[psm_steps$data_vanwestendorp$price == psm_steps$idp] - 0.05,
           label = paste("IDP: ", psm_steps$idp)) +
  coord_cartesian(xlim = c(10, 20)) +
  theme_minimal()
```

The Indifference Price Point (IDP) in this plot is not at the actual intersection of the "not cheap" and the "not expensive" curve. This is due to the data structure: We have sparse information for prices between 14 and 15. This can be seen in the rug plot close to the x-axis: Each vertical line indicates a price for which we have information in our data - however, there is a considerable gap around 14-15 (caused by the initial data manipulation that added price steps). The combination of price steps (as can be seen in the steep slope of the curves in this area) and sparse data points creates this problem. If the *interpolate* argument is set to *FALSE*, the *psm_analysis* function uses the first price for which the "not expensive" curve is higher than the "not cheap" curve, which is the case for a price of `r psm_steps$idp` in our example.

A larger sample size is usually sufficient to overcome this issue. However, this might not always be possible. Therefore, the *psm_analysis* function has an optional argument which helps to interpolate the pricing curve in cases of sparse information.

## Using the *interpolate* Argument

The *psm_analysis* function includes an optional argument *interpolate*. If this argument is set to *TRUE*, the function uses linear interpolation to estimate the price curves between the price points that are provided by the respondents. Consider an easy example: If the "not cheap" function is at 20% for a price of 5.99 and at 21% for a price of 6.01, the function would estimate that it is at 20.5% for a price of 6.00. The function uses fixed price steps of 0.01 units between the smallest and the largest price for which the respondents provide information.^[This might not always be practical, for instance in currencies where such a subdivision does not exist (such as the Chilean peso or the Tanzanian shilling).]

Below are the results for the same data when the *interpolate* function is set to *TRUE*.

```{r, fig.width=7, fig.height=4.5}
# Running the analysis with interpolation
psm_interpolated <- psm_analysis(toocheap = toocheap,
                                 cheap = cheap_alt,
                                 expensive = expensive_alt,
                                 tooexpensive = tooexpensive,
                                 validate = TRUE,
                                 interpolate = TRUE)

# Plot with interpolated price steps

psmplot_interpolated <-  ggplot(data = psm_interpolated$data_vanwestendorp, aes(x = price)) +
  geom_line(aes(y = ecdf_not_cheap, # line: not cheap
                colour = "not cheap"),
            size = 1) +
  geom_line(aes(y = ecdf_not_expensive, # line: not expensive
                colour = "not expensive"),
            size = 1) +
  geom_rug() +
  annotate(geom = "point", # Indifference Price Point (intersection of "cheap" and "expensive")
           x = psm_interpolated$idp,
           y = psm_interpolated$data_vanwestendorp$ecdf_not_expensive[psm_interpolated$data_vanwestendorp$price == psm_interpolated$idp],
           size = 5,
           shape = 18,
           colour = "#D55E00")

psmplot_interpolated +
  labs(x = "Price",
       y = "Share of Respondents (0-1)",
       title = "Price Sensitivity Meter Plot (with interpolation)")  +
  scale_colour_manual(name = "Legend",
                      values = c("not cheap" = "#009E73",
                                 "not expensive" = "#D55E00")) +
  annotate(geom = "text", # Label of Indifference Price Point
           x = psm_interpolated$idp + 1,
           y = psm_interpolated$data_vanwestendorp$ecdf_not_expensive[psm_interpolated$data_vanwestendorp$price == psm_interpolated$idp],
           label = paste("IDP: ", psm_interpolated$idp)) +
  coord_cartesian(xlim = c(10, 20)) +
  theme_minimal()
```

The rug plot demonstrates now that the data includes information for more pricing steps due to the interpolation.

The following plot demonstrates the close relationship between the original data and the interpolated data: The thicker semi-transparent line shows the interpolated values of the "not cheap" and "not expensive" curves. The thinner solid lines are the original data.

```{r, fig.width=7, fig.height=4.5, echo = FALSE}
psmplot_comparison <-  ggplot(data = psm_interpolated$data_vanwestendorp, aes(x = price)) +
  geom_line(aes(y = ecdf_not_cheap, # line: not cheap, interpolated
                colour = "not cheap"),
            size = 4, alpha = 0.25) +
  geom_line(aes(y = ecdf_not_expensive, # line: not expensive, interpolated
                colour = "not expensive"),
            size = 4, alpha = 0.25) +
  geom_line(data = psm_steps$data_vanwestendorp, # line: not cheap, not interpolated
            aes(y = ecdf_not_cheap, 
                colour = "not cheap"),
            size = 1) +
  geom_line(data = psm_steps$data_vanwestendorp, # line: not expensive, not interpolated
            aes(y = ecdf_not_expensive, 
                colour = "not expensive"),
            size = 1)

psmplot_comparison +
  labs(x = "Price",
       y = "Share of Respondents (0-1)",
       title = "Price Sensitivity Meter Plot: with vs. without interpolation")  +
  scale_colour_manual(name = "Legend",
                      values = c("not cheap" = "#009E73",
                                 "not expensive" = "#D55E00")) +
  coord_cartesian(xlim = c(10, 20)) +
  theme_minimal()
```
